---
title: "HW1"
output: html_document
date: "2023-01-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reproducibility component: 10 points.
1. (22 points total, equally weighted) The data set rnf6080.dat records hourly rainfall at a certain location in Canada, every day from 1960 to 1980.
a. Load the data set into R and make it a data frame called rain.df. What command did you use?
```{r}
read.table("rnf6080.dat")
```

b. How many rows and columns does rain.df have? How do you know? (If there are not 5070 rows and 27 columns, you did something wrong in the first part of the problem.)
```{r}

```

c. What command would you use to get the names of the columns of rain.df? What are those names?
```{r}

```

d. What command would you use to get the value at row 2, column 4? What is the value?
```{r}

```

e. What command would you use to display the whole second row? What is the content of that row?
```{r}

```

f. What does the following command do?
```{r}
names(rain.df) <- c("year","month","day",seq(0,23))
```

g. Create a new column called daily, which is the sum of the 24 hourly columns.
```{r}

```

h. Give the command you would use to create a histogram of the daily rainfall amounts. Please make sure to attach your figures in your .pdf report.
```{r}

```

i. Explain why that histogram above cannot possibly be right.
```{r}

```

j. Give the command you would use to fix the data frame.
```{r}

```

k. Create a corrected histogram and again include it as part of your submitted report. Explain why it is
more reasonable than the previous histogram.
```{r}

```

# Data types
2. (9 points, equally weighted) Make sure your answers to different parts of this problem are compatible
with each other.
a. For each of the following commands, either explain why they should be errors, or explain the nonerroneous
result.
```{r}
x <- c("5","12","7")
max(x)
sort(x)
sum(x)
```


b. For the next two commands, either explain their results, or why they should produce errors.
```{r}
y <- c("5",7,12)
y[2] + y[3]
```


c. For the next two commands, either explain their results, or why they should produce errors.
```{r}
z <- data.frame(z1="5",z2=7,z3=12)
z[1,2] + z[1,3]
```


3. (3 pts, equally weighted).
a.) What is the point of reproducible code?


b.) Given an example of why making your code reproducible is important for you to know in this class and
moving forward.

c.) On a scale of 1 (easy) – 10 (hard), how hard was this assignment. If this assignment was hard (> 5),
please state in one sentence what you struggled with.
